Data Specialization for a Staged Lambda Calculus

### 2015-02-26 paper/splitting.tex
Hacking on 4.0.  Trying to define the requirements of splitting more intuitively.

More precisely, for any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
$e$ splits into a syntactically separated program $\mathcal{P}$ 
(via the new relation $e \splittwosym \mathcal{P}$) which
can also be reduced to $q$ (via the new relation $\mathcal{P} \sepredtwosym q$).
By ``syntactically separated'' we mean that $\mathcal{P}$ 
has the form $\pipeS p l r$, where  
$p$ (called the {\em precomputation}) is a monostage term representing all the
first-stage computations in $e$, 
and $l.r$ (called the {\em resumer}) is a monostage term representing all the
second-stage computations in $e$.  
$\sepredtwosym$ is defined by a single rule, which evaluates the precomputation and substitutes the resulting value in for $l$ in the resumer:
\[
\infer{\sepredtwo {\pipeS p l r} {[b/l]r}} {\reduce p b}
\]
We call $b$ the {\em boundary value} resulting from the split since it represents the data communicated between the first and second stage computations.
A summary of these operations and the relationship between them is provided on the right side of \ref{fig:splittingSummary}.

Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via \texttt{prev}), splitting is also defined on 
\bbonem\ terms. %we must also provide a way to split those as well.
This is given by the new relation $e \splitonesym \mathcal{P}$,
where $\mathcal{P}$ must have the form $\pipeM c l r$ where  
$c$ (called the {\em combined term}) is a monostage term encoding all the first-stage computations in $e$, 
and $l.r$ (again called the {\em resumer}) is a monostage term encoding all the
second-stage computations in $e$.

Similar to above, we also define a $\sepredtwosym$ relation 
which can be used to evaluate these seperated world \bbonep\ terms.
One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
For example, consider the term
\begin{lstlisting}
(next {1+2}, grnd{3+4})
\end{lstlisting}
which reduces (via $\redonesym$) to the residual table and partial value
\begin{lstlisting}
[yhat |-> 1+2] (next {yhat}, grnd{7})
\end{lstlisting}
Since this output has interleaved stages, theres no way a syntactically separated term could directly reduce to it.

We resolve this mismatch by introducing a translation, called {\em masking} and written $\rtab \xi v \vsplito \mathcal{V}$,
which converts residual tables and partial values---like those in the example above---into a syntactically separated version $\mathcal{V}$.
As before, we enforce the ``syntactically separated'' property by defining $\mathcal{V}$ to have the form $\mval i q$ (called a {\em masked value}) where 
$i$ is a monostage value representing the first stage part of $v$,
and $q$ is a monostage term representing all second-stage computations in $\xi$ and $v$.
Note that masking is a technical device used to state the correctness of splitting at \bbonem; 
it is not part of the splitting algorithm itself.

>>>>

More precisely, for any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
$e$ splits into a syntactically separated program $\mathcal{P}$ 
(via the new relation $e \splittwosym \mathcal{P}$) which is
reduced to a residual $q'$ (via the new relation $\mathcal{P} \sepredtwosym q'$), which is the same as $q$ up to an equivalence.
By ``syntactically separated'' we mean that $\mathcal{P}$ 
has the form $\pipeS p l r$, where  
$p$ (called the {\em precomputation}) is a monostage term representing all the
first-stage computations in $e$, 
and $l.r$ (called the {\em resumer}) is a monostage term representing all the
second-stage computations in $e$.  
$\sepredtwosym$ is defined by a single rule, which evaluates the precomputation and substitutes the resulting value in for $l$ in the resumer:
\[
\infer{\sepredtwo {\pipeS p l r} {[b/l]r}} {\reduce p b}
\]
We call $b$ the {\em boundary value} resulting from the split since it represents the data communicated between the first and second stage computations.
A summary of these operations and the relationship between them is provided on the right side of \ref{fig:splittingSummary}.

Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via \texttt{prev}), splitting is also defined on 
\bbonem\ terms.
However, defining correctness requirements for splitting \bbonem\ terms is
challenging because unlike \bbtwo\ terms (which reduce to monostage
residuals), \bbonem\ terms reduce to partial values and
residual tables that represent computations at both stages.
%One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
For example, the \bbonem\ term:
\begin{lstlisting}
(next {1+2}, grnd{3+4})
\end{lstlisting}
reduces (via $\redonesym$) to the residual table and partial value:
\begin{lstlisting}
[yhat |-> 1+2] (next {yhat}, grnd{7})
\end{lstlisting}

To resolve the mismatch between the multi-stage computations
represented by residual tables and partial values and the desired monostaged
outputs of splitting,
% Since this output has interleaved stages, theres no way a syntactically separated term could directly reduce to it.
we define a translation, called {\em masking}, written $\rtab \xi v \vsplito \mathcal{V}$,
which converts residual tables and partial values into a syntactically separated {\em masked value} of the form $\mval i q$.
%version $\mathcal{V}$.
%As before, we enforce the ``syntactically separated'' property by defining $\mathcal{V}$ to have the form $\mval i q$ (called a {\em masked value})
The first component, $i$ of the masked value holds the result of all
first stage computations in $v$.  The second component, $q$ is a
residual representing all second stage computations in $\rtab \xi v$.

%where 
%$i$ is a monostage value representing the first stage part of $v$,
%and $q$ is a monostage term representing all second-stage computations in $\xi$ and $v$.

%Note that masking is a technical device used to state the correctness of splitting at \bbonem; 
%it is not part of the splitting algorithm itself.

Given this definition of masking, the requirements for splitting
\bbonem\ terms are as follows. For any $\colmix e A$ which reduces to
$\rtab \xi v$ (via $\redonesym$), $\rtab \xi v$ masks to $\mval i q$
(via $\vsplito$), $e$ splits into a syntactically seperated program
$\mathcal{P}$ of the form $\pipeM c l r$ (via the new relation $e
\splitonesym \mathcal{P}$) which is reduced to a masked value $\mval i
{q'}$ (via the new relation $\sepredtwosym$) where $q'$ is the same as
$q$ up to an equivalence.

%Given the definition of masking, \bbonem\ terms split via the new
%relation $e \splitonesym \mathcal{P}$, where $\mathcal{P}$ must have
%the form $\pipeM c l r$
where $c$ (called the {\em combined term}) is
a monostage term encoding all the first-stage computations in $e$, and
$l.r$ (again called the {\em resumer}) is a monostage term encoding
all the second-stage computations in $e$.

Similar to above, we also define a $\sepredtwosym$ relation 
which can be used to evaluate these seperated world \bbonep\ terms.
